---
title: "Stack benchmarks"
author: "Barret Schloerke"
date: "4/8/2021"
output: html_document
---


# Which stack produces unique environments the fastest?

Finding the unique environments is required for `tagQuery()`'s internal methods.  However, `unique()` has `O(n^2)` complexity. At some point it performs poorly.

## Goal

Find a stack that performs well given different stack sizes (`n`).

## Setup

If unique keys needs to be managed, `rlang::sexp_address(env)` will be used.  This method is MUCH faster than `format.default(env)` and achieves the same goal of providing a unique key for each environment.


```{r bench_addr}
library(fastmap)
library(rlang)
library(magrittr)
library(ggplot2)

# iterate over items in `x` with function `f`
walk <- function(.x, .f, ...) {
  for (i in seq_along(.x)) {
    .f(.x[[i]], ...)
  }
  NULL
}

# Show that `rlang:::sexp_address(env)` is faster than `format.default(env)`
local({
  rlang_addr <- rlang:::sexp_address
  format_addr <- format.default
  env <- new.env()
  bench::mark(check = FALSE,
    rlang_addr(env),
    format_addr(env)
  )
})
```


# Stacks

### Standard stack

Very low overhead. Performs poorly as `n` becomes large due to `unique()` call.

```{r standard_stack}
standard_stack <- function() {
  stack <- faststack()
  list(
    push = stack$push,
    as_list = stack$as_list,
    unique_list = function() {
      unique(stack$as_list())
    }
  )
}
```

### Unique stack

Only allows unique elements into the stack. Leverages a `fastmap()` to know which elements have been entered into the stack.

```{r uni_stack}
uni_stack <- function() {
  addr <- rlang:::sexp_address
  map <- fastmap()
  stack <- faststack()
  list(
    push = function(env) {
      key <- addr(env)
      if (!map$has(key)) {
        map$set(key, TRUE)
        stack$push(env)
      }
    },
    unique_list = stack$as_list
  )
}
```

### Hybrid stack

Performs like a standard stack until it hits 1000 elements. At 1000 elements, the internal stack is converted to a `uni_stack()`.

While there is overhead at the 1000 mark, the time penalty should be recovered as `n` becomes larger as it will behave like a `uni_stack()`.

```{r hybrid_stack}
hybrid_stack <- function() {
  stack <- standard_stack()
  count <- 0
  list(
    push = function(env) {
      count <<- count + 1
      if (count == 1000) {
        # convert the current stack to a `uni_stack()`
        new_stack <- uni_stack()
        walk(stack$as_list(), new_stack$push)
        stack <<- new_stack
      }
      stack$push(env)
    },
    unique_list = function() {stack$unique_list()}
  )
}
```


# Testing

We are interested in where the hybrid stack should be converted from a standard stack to a uni stack.  Testing needs to be execute both the ingestion of environments and retrieval of the unique list.

### Benchmark

Let's run a benchmark 10 times over increasing values of `n` for all three stacks.

```{r bench}
info <-
  lapply(1:5, function(i) {
    lapply(
      c(10,25,50,75,100, 250,500,750,1000,2500,5000,7500),
      function(n) {
        cat("i: ", i, "; n: ", n, "\n", sep = "")
        envs <- lapply(seq_len(n), function(i) {
          new.env(parent = emptyenv())
        })

        populate <- function(set) {
          walk(envs, function(env) {
            set$push(env)
          })
          set
        }

        ret <-
          bench::mark(
            populate(standard_stack())$unique_list(),
            populate(uni_stack())$unique_list(),
            populate(hybrid_stack())$unique_list()
          )
        ret$n <- n
        ret$i <- i
        ret
      }
    )
  })
```

### Data

Quick inspection of the data...

```{r inspect_data}
dt <- info %>%
  unlist(recursive = FALSE) %>%
  dplyr::bind_rows() %>%
  dplyr::select(expression, median, n) %>%
  dplyr::mutate(
    run = purrr::map_chr(expression, as.character)) %>%
  dplyr::select(-expression) %>%
  tibble::glimpse()

dt_avg <-
  dt %>%
  dplyr::group_by(run, n) %>%
  dplyr::summarise(median = median(median)) %>%
  tibble::glimpse()
```

# Results

Switching from a standard stack to a unique stack at ~ 500 seems to make sense as that is where the `standard_stack()` line and the `uni_stack()` line cross. 

```{r plot}
ggplot(dt, aes(n, median * 100000, color = run)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  geom_line(data = dt_avg) +
  geom_vline(xintercept = 1000) +
  labs(y = "median bench time (micro sec)")
```


```{r}
ratios <- 
  dt_avg %>% 
  dplyr::ungroup() %>% 
  dplyr::filter(n %in% c(10, 5000)) %>% 
  tidyr::pivot_wider(run, names_from = n, names_prefix = "n_", values_from = median) %>%
  dplyr::mutate(
    ratio_10 = as.numeric(n_10) / as.numeric(n_10)[1], 
    ratio_5000 = as.numeric(n_5000) / as.numeric(n_5000)[1]
  ) %>% 
  print()
```

The hybrid stack is `r ratios$ratio_10[3]` times faster than `uni_stack()` when `n` is 10.

The hybrid stack is `r ratios$ratio_5000[2]` times faster than `standard_stack()` when `n` is 5000.
